\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage[hidelinks]{hyperref}
\usepackage{enumitem}
\setlist{nosep}

\title{Recommender Model Review V2 (Industry-heavy)}
\author{Kelly (OpenClaw) \\\textit{Prepared for Lewis}}
\date{\today}

\begin{document}
\maketitle

\section*{Scope}
This V2 review enriches our previous set of industry-oriented recommender model papers (OneRec V1/V2, HSTU(+Ultra), PLUM, OneTrans) and extends the categorization by adding more industry papers that highlight recurring systems/serving constraints.

\section{Updated categorization (V2 taxonomy)}
\begin{itemize}
  \item \textbf{A. End-to-end generative recommendation / unified retrieval+ranking} (OneRec V1/V2, OneMall): single generative policy produces semantic IDs; RL alignment.
  \item \textbf{B. Generative recommenders as sequential transducers (foundation-style scaling)} (HSTU / GRs, Ultra-HSTU): ranking/retrieval as transduction; scaling laws + long-context efficiency.
  \item \textbf{C. Industrial ranking Transformers unifying sequence modeling + feature interaction} (OneTrans): unify tokenization; mixed parameterization; pyramid pruning + KV caching.
  \item \textbf{D. Ultra-long history via offline embeddings (feature pipeline as product)} (DV365): encode 10k--70k history offline into embeddings shared across many downstream models.
  \item \textbf{E. End-to-end ultra-long sequence Transformers with GPU-centric engineering} (LONGER): global tokens, token merge, hybrid attention + mixed precision/recompute/KV cache/sync train-serve.
  \item \textbf{F. Lifelong sequences inside CTR ranking + serving optimization} (TransAct V2): lifelong sequences with next-action loss and web-scale serving optimizations.
  \item \textbf{G. Infra-first systems for online training + embedding tables under concept drift} (Monolith): collisionless embedding tables + online training; reliability vs real-time tradeoffs.
\end{itemize}

\section{What new challenges industry papers emphasize (cross-paper themes)}
\begin{itemize}
  \item \textbf{Compute allocation matters more than raw model size:} e.g., OneRec-V2 motivates shifting FLOPs from context encoding to decision-critical generation.
  \item \textbf{Ultra-long history is valuable, but serving it is the constraint:} DV365 (offline embeddings), TransAct V2 (serving constraints), LONGER (GPU-efficient long-context).
  \item \textbf{Two-stage ``top-k history'' retrieval creates inconsistency:} LONGER highlights upstream/downstream mismatch in common practice.
  \item \textbf{Multi-surface production makes shared representations valuable:} DV365 shares embeddings across 15 models; OneMall/OneRec unify scenarios.
  \item \textbf{Unified backbones reuse LLM infra optimizations:} OneTrans emphasizes KV caching, memory-efficient attention, mixed precision.
  \item \textbf{Alignment is hard:} OneRec family focuses on RL alignment and real-feedback shaping; proxy rewards can be hacked.
  \item \textbf{Concept drift forces online learning + embedding-table engineering:} Monolith centers sparse/dynamic features and real-time training.
\end{itemize}

\section{Paper set (core + new for V2)}
\subsection*{Core (reviewed together)}
\begin{itemize}
  \item OneRec-V1 (Kuaishou): \texttt{papers/Kuaishou\_OneRec\_V1/}
  \item OneRec-V2: \texttt{papers/OneRecTeam\_OneRecV2/}
  \item HSTU (Meta): \texttt{papers/meta\_hstu/}
  \item Ultra-HSTU: \texttt{papers/meta\_ultra\_hstu/}
  \item PLUM (Google/YouTube): \texttt{papers/google\_plum/}
  \item OneTrans (ByteDance): \texttt{papers/bytedance\_onetrans/}
\end{itemize}

\subsection*{New industry papers included for V2 expansion}
\begin{itemize}
  \item DV365 (Instagram): arXiv:2506.00450
  \item LONGER (ByteDance): arXiv:2505.04421
  \item TransAct V2 (Pinterest): arXiv:2506.02267
  \item Monolith (ByteDance): arXiv:2209.07663
\end{itemize}

\section{Next steps}
\begin{enumerate}
  \item Extract and consolidate the \textbf{References} sections from the core papers, and build a reference graph (themes: long-seq, SIDs/tokenization, RL alignment, infra/embedding tables).
  \item Add 5--10 more industry papers following the citation graph (e.g., SIM/TWIN/DIN/TransAct/LONGER/Monolith-related).
  \item Expand this V2 into a full evidence-backed review (quotes + tables) and publish as a PR (PDF + \LaTeX source).
\end{enumerate}

\end{document}
